Case Study: ChIP-seq Differential Peak Calling (pre-defined regions)

### 2018-03-08 datasets/ChIP-seq/ChIPseq-promoters.Rmd
Start working on csaw, diffbind analysis

---
title: "Case Study: ChIP-seq Differential Peak Calling (predefined regions)"
author: "Mingxiang Teng"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

In differential peak calling, classical pipelines often begin by first calling peaks
in samples, followed by performing differential testing using one of a few methods,
e.g. _csaw_ or _DiffBind_. However, with these approaches, testing is only performed on a
filtered set of genomic regions, namely those with significance in at least one condition,
resulting in a potentially limited number of null tests cases. An analysis using this
classical approach is performed in the accompanying `ChIPseq-peaks.Rmd` case study.

To avoid this limitation, in this case study we examine ChIP-seq data of the histone
H3K4Me3, an active marker of gene expression, and focus our analysis on promoter regions
of all genes for this histone marker. By doing so, we avoid limiting the set of tests
to only regions significant in at least one sample, while also keeping most regions of
interest for this histone marker.

As in `ChIPseq-peaks.Rmd`, we test for differential peaks using publicly available ChIP-seq
data for histone H3K4Me3 from two cell lines (K562 and Gm12878). Data are obtained from the
ENCODE Project. Differential peaks are tested using DESeq2.

@@

@@

@@

We count sequencing reads for all promoter regions here, and test for differential peaks using these counts.

```{r count, results='hide', message=FALSE, warning=FALSE}
deseq_res  <- file.path(resdir, "h3k4me3-promotors-deseq.rda")
if (file.exists(deseq_res)) {
    load(deseq_res)
} else {
    ## count reads in each promotor region
    anno <- data.frame(GeneID = seq_len(length(prom)),
                       Chr = seqnames(prom),
                       Start = start(prom),
                       End = end(prom),
                       Strand = strand(prom))
    rc <- featureCounts(files = bamfiles,
                        annot.ext = anno,
                        allowMultiOverlap = TRUE,
                        minOverlap = 50,
                        readExtension3 = 150,
                        ignoreDup = TRUE)$counts

    ## perform differential testing
    dds <- DESeqDataSetFromMatrix(countData = rc,
                                  colData = as.data.frame(meta),
                                  design= ~ lab + cellline)
    dds <- DESeq(dds, fitType = "mean")
    dat <- as.data.frame(results(dds))
    colnames(dat) <- c('ind_covariate', 'effect_size', 'SE',
                       'stat', 'pval', 'padj')

    ## compute test-statistic from p-value for ASH
    dat <- dplyr::mutate(dat,
                         test_statistic = qnorm(exp(log(pval)-log(2)),
                                                lower.tail=FALSE),
                         test_statistic = test_statistic * sign(effect_size))

    save(deseq, dat, file = deseq_res)
}

## remove regions with effect sizes = NA
dat <- dat[!is.na(dat[, 2]), ]
```

@@

Here, we can see that the covariant variable is actually informative.

@@

```{r, chipseq-sb, results="hide", message=FALSE}
bench_res  <- file.path(resdir, "h3k4me3-promotors-sb.rda")
if (file.exists(bench_res)) {
    load(bench_res)
} else {
    sb <- buildBench(bd, data = dat, ftCols = c("ind_covariate"),
                     parallel = TRUE, BPPARAM = multicoreParam)
    save(sb, file = bench_res)
}
```

>>>>

---
title: "Case Study: ChIP-seq Differential Peak Calling (pre-defined regions)"
author: "Mingxiang Teng, Patrick Kimes"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

In this case study, we perform differential peak calling on ChIP-seq data of the histone
H3K4Me3 for samples from two cell lines (K562 and Gm12878) using publicly available data
generated by the ENCODE Project. The same data set is used for all ChIP-seq differential
testing case studies.

While several approaches have been proposed for differential peak calling, in this case
study we examine a simple approach of testing for differential peaks in the promoter 
regions of all genes. While using pre-defined regions can reduce the power of the test to
detect differences in non-promoter regions, since H3K4Me3 is an active marker of gene
expression, restricting tests to promoter regions is not unreasonable. Differential peaks
are tested using DESeq2.

@@

## intermediary files we create below
count_file <- file.path(resdir, "h3k4me3-promotors-counts.rda")
result_file <- file.path(resdir, "h3k4me3-promotors-results.rda")
bench_file <- file.path(resdir, "h3k4me3-promotors-benchmark.rda")

@@

## Read Counting

We count reads for all promoter regions.

```{r count, results='hide', message=FALSE, warning=FALSE}
if (file.exists(count_file)) {
    load(count_file)
} else {
    anno <- data.frame(GeneID = seq_len(length(prom)),
                       Chr = seqnames(prom),
                       Start = start(prom),
                       End = end(prom),
                       Strand = strand(prom))
    rc <- featureCounts(files = bamfiles,
                        annot.ext = anno,
                        allowMultiOverlap = TRUE,
                        minOverlap = 50,
                        readExtension3 = 150,
                        ignoreDup = TRUE)$counts
    save(rc, file = count_file)
}
```

@@

We test for differential peaks at each of the .

```{r count, results='hide', message=FALSE, warning=FALSE}
if (file.exists(result_file)) {
    load(result_file)
} else {
    dds <- DESeqDataSetFromMatrix(countData = rc,
                                  colData = as.data.frame(meta),
                                  design = ~ lab + cellline)
    dds <- DESeq(dds, fitType = "mean")
    dat <- as.data.frame(results(dds))
    colnames(dat) <- c('ind_covariate', 'effect_size', 'SE',
                       'stat', 'pval', 'padj')
    ## compute test-statistic from p-value for ASH
    dat <- dplyr::mutate(dat,
                         test_statistic = qnorm(exp(log(pval)-log(2)),
                                                lower.tail=FALSE),
                         test_statistic = test_statistic * sign(effect_size))
    ## remove regions with effect sizes = NA
    dat <- dat[!is.na(dat$effect_size), ]
    save(dat, file = result_file)
}
```

@@

Here, we can check diagnostic plots to assess whether candidate covariates are actually informative.

@@

```{r, chipseq-sb, results="hide", message=FALSE}
if (file.exists(bench_file)) {
    load(bench_file)
} else {
    sb <- buildBench(bd, data = dat, ftCols = c("ind_covariate"),
                     parallel = TRUE, BPPARAM = multicoreParam)
    save(sb, file = bench_file)
}
```
